---
title: "WS temporais"
author: "Cesar Cordeiro"
date: "8/10/2020"
output: html_document
---

```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = TRUE, fig.cap = "", fig.path = "Plot")

```



```{r read_files, echo=FALSE, warning=FALSE}

source("script/sumario.R")

```


```{r habitats}

dados %>% 
  filter(peld == 'PELD') %>% 
  select(nome_programa, habitat) %>%
  mutate(habitat = plyr::mapvalues(habitat, from = 'estuario_marinho', to = 'estuario')) %>% 
  distinct() %>% 
  bind_rows(data.frame(nome_programa = "PELD-RECA", habitat = c("recife_rochoso", "entremares_rochoso"))) %>% 
  aggregate(nome_programa ~ habitat, ., function(x) length(unique(x)))

```


# Title: How long-term monitoring programs has helped marine conservation in the Brazilian coast? 


### Abstract (< 200 words) and graphical abstract


### Highlights (< 5 bullet points)

* Most of marine long-term ecological monitoring sites are recent in Brazil *
* Scientific production associated with PELD sites increased exponentially with program age *
* Connectivity among PELD sites is low *
* Financing is a bottleneck for long-term ecological monitoring initiatives *


## Introduction (600 words)

Monitoring biodiversity can provide valuable information for management, sustainability and our own survivorship face the increasing changes in ecosystems’ functioning and degradation of ecological services (Lindenmayer & Likens, 2009). Such biodiversity supervision is key to act before and while changes are still manageable. Monitoring is also about collaboration among social and political entities to help decision making process of resource management (Magnusson et al., 2013). Lastly, it also depends in optimal planning of data management to maximize usage and ensure its longevity (Magnusson et al., 2013).
Long-term ecological monitoring are essential tools to raise red flags on species’ decline, species distributional changes, increase in functional vulnerability and risk of species extinction (Gaiser et al., 2020; Lindenmayer & Likens, 2009) helping to track back and pinpoint general changes to ecosystems. On these efforts, special attention should focus on tropical regions where climatic changes may deeply compromise biodiversity because species mainly live close to their physiological optima (Sheldon, 2019; Worm & Lotze, 2016). Globally, monitoring efforts are mostly associated to northern hemisphere biomes following historical background of wealthier nations (Mirtl et al., 2018).
The PELD program (Programa Ecológico de Longa Duração in portuguese) is a national effort funded by the Brazilian government that seeks to conduct research at large temporal scales and large spatial scales (Barbosa, 2010). This program was inspired by the LTER (Long-term Ecological Research) network created by the NSF (National Science Foundation) in the United States, and started in 1997 in Brazil coordinated by the CNPq (Conselho Nacional de Desenvolvimento Científico e Tecnológico in portuguese) (Barbosa, 2010). There are 35 PELD sites in the Brazilian territory, and nine of those sites encompass marine or estuarine habitats (Brito et al., 2020), which so far have been working independently with negligible direct cooperation.
While we enter in the Ocean Decade where clean, accessible and resilient oceans are key to achieve sustainable objective goals (Heymans et al., 2020; Visbeck, 2018), Brazil’s role in monitoring the South Atlantic Ocean is crucial since its territorial waters represent large part of it. Our objective in this workshop was to join all PELD sites monitoring marine and estuarine ecosystems and compile the main information and products generated by those monitoring nuclei to make a diagnosis of the actual state of those monitoring sites and build a working basis for future conjunct efforts and collaboration. This manuscript describes the characteristics of those PELD sites indicating gaps and future directions for the biodiversity monitoring in the Brazilian coast.
This research is one of the products of the first “Workshop for integration of time series from marine ecosystems in Brazil – Long-Term Ecological Program (PELD)” supported by the Fundação de Amparo a Pesquisa do Rio de Janeiro (FAPERJ). This effort sought to bring together research groups operating on the Brazilian coast is also in line with the objectives of the decade of the oceans associated with help identifying knowledge gaps, potential effects of climate change on the functioning of ecosystems and increasing the visibility and availability of information about the oceans. Also, this initiative helps to strengthen the long-term monitoring programs, especially the PELD, financed by the CNPq which had the presence of a representative at the event and was able to indicate the current scenario and involvement of the institution with efforts in the research of climate change. Thus, we investigated the main characteristics of long-term monitoring programs established in Brazilian coastal environments to investigate their main characteristics concerning scientific production, financing and involved human resources as well as the spatial and temporal scales covered by those initiatives. From that picture we expected to evaluate gaps and strengths to help framing future directions on the path of coastal environment monitoring.


## Material and Methods (max 600 words)

The event was attended by 28 researchers, from 16 universities and research institutes from eight states and the Federal District. Researchers attending the event were associated with six sites of PELD program (table 1). We also had the contribution of researchers monitoring other sites supported mostly by public funds but not already stablished as PELD sites. Most of those non-PELD sites have been monitored by many years relying on individual efforts of a single or a small group of scientists. This data was not included in this manuscript because it represents individual efforts and may be underestimated needing further detailing. Also, well known long-term monitoring programs supported by oil companies and other mixed-funds sources were not treated here. Thus, only the data from research groups monitoring PELD sites were included in the analyses. The dynamics of the workshop was divided into three stages: identification of the temporal and spatial scales of the monitoring programs, evaluation of possible approaches to the analysis of time series and forwarding of actions for the execution of products from the existing data. The following indicates the results from the workshop first stage and was based on information collected from coordinators and collaborators of PELD monitoring programs.
	Data from each project was obtained by direct interview with coordinators and complemented by answering digital forms (Google Forms®) sent after the meeting. Questions from used in forms and variables evaluated can be found on the supplementary material (table 1S).



```{r map, echo=FALSE, fig.cap = "Figure 1. Map with monitored sites, length of time series, habitats monitored and indicators"}

# https://www.r-spatial.org/r/2018/10/25/ggplot2-sf.html

### MAP

# MPA
formulario %>% 
  filter(PELD == "Sim") %>%  
  dplyr::select(nome_programa, apenas_uc) %>%
  distinct()

# sites
mapa <- ggplot(data = world) + 
    geom_sf(fill = "antiquewhite", color = "white") + 
    annotate(geom = "text", x = -90, y = 26, label = "label", fontface = "italic", color = "grey22", size = 6) + 
    annotation_north_arrow(location = "br", which_north = "true", 
                           pad_x = unit(0.2, "cm"), pad_y = unit(0.2, "cm"), 
                           style = north_arrow_fancy_orienteering,
                           height = unit(1, "cm"),
                           width = unit(1, "cm")) + 
    coord_sf(xlim = c(-60, -15), ylim = c(-40, 10), expand = FALSE) + 
    geom_polygon(data = fortify(priori_ll), aes(x = long, y = lat, group = group), fill = 'red', alpha = 0.9) +
    labs(x="longitude", y="latitude", title = "(A)") +
    theme_classic() +
    theme(panel.background = element_rect(fill = "aliceblue"),
          legend.title = element_blank(),
          text = element_text(size = 16),
          legend.position = "bottom") +
    geom_point(data = subset(pontos, peld == "PELD"), aes(x = lon, y = lat), shape = 19, size = 2) +
    geom_text_repel(data = subset(pontos, peld == "PELD"), 
                    aes(x = lon, y = lat, label = nome_programa), size = 4,
                    alpha=0.7, segment.color = 'grey50', segment.size = 0.4, 
                    box.padding = unit(0.8, 'lines')) #+
    # geom_point(data = subset(pontos, peld == "non-PELD"), aes(x = lon, y = lat), shape = 3, size = 1) +
    # geom_text_repel(data = subset(pontos, peld == "non-PELD"), 
    #                 aes(x = lon, y = lat, label = nome_programa), size = 3, 
    #                 alpha=0.7, segment.color = 'grey50', segment.size = 0.4, 
    #                 box.padding = unit(0.8, 'lines'))
  

# monitoring time spam
anos <- ggplot(data.frame()) +
  geom_point() +
  xlim(1974, 2020) +
  geom_segment(aes(x = 2003, y = "ILOC", xend = 2016, yend = 9)) + # ILOC
  geom_segment(aes(x = 2016, y = "ILOC", xend = 2020, yend = 9), color = "darkblue", size = 3) + # ILOC
  geom_segment(aes(x = 2010, y = "TAMS", xend = 2020, yend = 8)) + # TAMS +
    geom_segment(aes(x = 2016, y = "TAMS", xend = 2020, yend = 8), color = "darkblue", size = 3) + # TAMS
  geom_segment(aes(x = 2016, y = "CCAL", xend = 2020, yend = 7), color = "darkblue", size = 3) + # CCAL
  geom_segment(aes(x = 2006, y = "Abrolhos", xend = 2020, yend = 6), color = "darkblue", size = 3) + # Abrolhos
  geom_segment(aes(x = 2012, y = "HCES", xend = 2016, yend = 5)) + # HCES
  geom_segment(aes(x = 2016, y = "HCES", xend = 2020, yend = 5), color = "darkblue", size = 3) + # HCES
  geom_segment(aes(x = 2000, y = "RLaC", xend = 2020, yend = 4), color = "darkblue", size = 3) + # RLaC
  geom_segment(aes(x = 1974, y = "RECA", xend = 2016, yend = 3)) + # RECA
  geom_segment(aes(x = 2016, y = "RECA", xend = 2020, yend = 3), color = "darkblue", size = 3) + # RECA
  geom_segment(aes(x = 1997, y = "PEBG", xend = 2020, yend = 2), color = "darkblue", size = 3) + # PEBG
  geom_segment(aes(x = 1993, y = "ELPA", xend = 2020, yend = 1), color = "darkblue", size = 3) + # ELPA
  scale_y_discrete(limits=rev(c("ILOC", "TAMS", "CCAL", "Abrolhos", "HCES", "RLaC", "RECA", "PEBG", "ELPA"))) + 
  theme_classic() + 
  theme(#axis.text.y=element_blank(),
        text = element_text(size = 16),
        axis.text.x=element_text(angle=90, vjust = 0.5, hjust = 1)) +
  labs(x = "", y ="", title = "(B)")


# biotic variables monitored
biotic <- dados %>%
  filter(peld == "PELD") %>%
  select(nome_programa, grupos_biologicos) %>%
  bind_rows(data.frame(nome_programa = c("PELD-RECA"),
                       grupos_biologicos = c("sessile invertebrates", "aquatic vegetation", "others"))) %>%
  mutate(nome_programa = factor(nome_programa, levels = rev(c("PELD-ILOC", "PELD-TAMS", "PELD-CCAL", "PELD-Abrolhos", "PELD-HCES", "PELD-RLaC", "PELD-RECA", "PELD-PEBG", "PELD-ELPA")))) %>%
  mutate(grupos_biologicos = plyr::mapvalues(grupos_biologicos, from = c("fitoplancton", "zooplancton", "podutores-consumidores", "macrofauna_>0.5cm", "vegetacao-aquatica-submersa",                                                                      "peixes","ictioplancton","Tursiops truncatus", "bentos", "eDNA", "invertebrados", "estoque_carbono", 
                                                                         "liteira", "coral", "grapsus", "montastrea", "siderastrea", "ouricos", "bacterioplancton", 
                                                                         "meroplancton", "holoplancton", "Didelphis aurita", "Philander quica", "Cerradomis goytaca", 
                                                                         "Atlantorchestoidea brasiliensis", "Quadrivisio lutzi", "populacao-Clusia", "vegetacao_clusia-não-separada",
                                                                         "bentos_consolidado", "grama marinha", "Macroalgas", "produtores", "fontes de matéria orgânica", "ictiofauna",
                                                                         "vegetacao", "resources and socio-economics", "reef_accretion"), 
                                             to = c("plankton", "plankton", NA,"mobile invertebrates", "aquatic vegetation",
                                                    "fish","plankton","marine mammals", "aquatic vegetation", "others", "sessile invertebrates", "others", 
                                                    NA, "sessile invertebrates", "mobile invertebrates", "sessile invertebrates", "sessile invertebrates", "mobile invertebrates", "plankton", 
                                                    "plankton", "plankton", "terrestrial mammals", "terrestrial mammals", "terrestrial mammals", 
                                                    "mobile invertebrates", "terrestrial mammals", "coastal vegetation", "coastal vegetation", "sessile invertebrates",
                                                    "aquatic vegetation", "aquatic vegetation", "aquatic vegetation", "others",
                                                    "fish", "coastal vegetation", "socio-economics", "others"))) %>% 
  filter(!is.na(grupos_biologicos)) %>% 
  distinct() %>% 
   ggplot(aes(y = nome_programa, x = grupos_biologicos)) +
  geom_point(size = 5, color = "darkgreen") +
  theme_classic() + 
  theme(axis.text.y = element_blank(),
        text = element_text(size = 16),
    axis.text.x = element_text(angle=90, vjust = 0.5, hjust = 1),
    legend.position = "") +
    labs(x = "", y ="", title = "(C)") +
    scale_x_discrete(limits=c("fish", "mobile invertebrates", "aquatic vegetation", "sessile invertebrates", "plankton",
                              "socio-economics", "coastal vegetation", "others", "marine mammals", "terrestrial mammals"))

# plot
mapa + anos + biotic + plot_layout(nrow=1, widths = c(6, 1.5, 3), heights = 4) # save in dimensions 6x12 pol

```
Figure 1. Map with monitored sites (A), length of time series (B), and general indicators monitored (C) at each site. Polygons in red (A) indicate priority areas for conservation according to Margris et al. (2021). ‘others’ = may include isotopes, carbon stock, reef accretion and primary production.


## Results (max. 4 figs or tables and 700 words)

Programs and local efforts identified along the Brazilian coast included 21 sites along approximately 3200 km of coast, from Pernambuco to Rio Grande do Sul, with nine of those sites being monitored by PELD associated groups. Large gaps are found in the north and northeast regions (between Amapá and Rio Grande do Norte), but also at the southeast region (between São Paulo and Rio Grande do Sul). Habitats monitored by the investigated initiatives included mainly the subtidal realm, while terrestrial elements (coastal vegetation and fauna) were present only in two programs (HCES and RLaC). Monitored habitats included only shallow areas (< 50 m deep) both inshore and offshore, where reefs environments (biogenic and non-biogenic) were the most widespread habitat surveyed in six out of nine monitoring sites (table 1). The distribution of reef environments included both tropical and subtropical domains and all Brazilian oceanic islands.


```{r habitats, echo=FALSE}
habitat <- dcast(dados, local + lat_mean + lon_mean ~ habitat, value.var="habitat")
habitat1 <- cbind(habitat[,1:3], decostand(habitat[,4:ncol(habitat)], method = 'pa'))

leaflet() %>%
  addTiles() %>%
  addMinicharts(
    habitat1$lon, habitat1$lat,
    type = "polar-radius",
    chartdata = habitat1[, 5:11],
    #width = 20, transitionTime = 0 #60 * sqrt(prod2016$total) / sqrt(max(prod2016$total))
  )
```
Table 1. Long-term ecological monitoring sites from PELD in Brazilian coastal environments with respective habitats monitored, frequency of monitoring and level of protection.


Reef fish species were the biological group with the greatest spatial amplitude in monitoring, covering tropical and subtropical domains (figure 1). The aquatic vegetation and the mobile invertebrate fauna were monitored sparsely along the coast, and neritic components were mainly monitored at southern sites (figure 1).


```{r table, echo=FALSE}

# estudantes
students <- formulario %>% 
  dplyr::filter(PELD == "Sim") %>% 
  dplyr::select(nome_programa, n_estudantes) %>% #, n_pesquisadores, teses_etc, artigos, outro_produto
  unique() %>%
  mutate(n_estudantes = strsplit(as.character(n_estudantes), ";")) %>%
  unnest(cols = c(n_estudantes)) %>% 
  mutate(n_estudantes = trimws(.$n_estudantes)) %>% 
  separate(n_estudantes, c("categoria", "estudantes"), " = ") %>% 
  unnest(cols = c()) %>% 
  mutate(estudantes = as.numeric(ifelse(categoria %in% c(31, 4, 19, 50), categoria, estudantes))) %>% 
  group_by(nome_programa) %>% 
  summarise(students = sum(estudantes))

# teses
thesis <- formulario %>% 
  dplyr::filter(PELD == "Sim") %>% 
  dplyr::select(nome_programa, teses_etc) %>% #, n_pesquisadores, teses_etc, artigos, outro_produto
  unique() %>%
  mutate(teses_etc = strsplit(as.character(teses_etc), ";")) %>%
  unnest(cols = c(teses_etc)) %>% 
  mutate(teses_etc = trimws(.$teses_etc)) %>% 
  separate(teses_etc, c("categoria", "estudantes"), " = ") %>% 
  mutate(estudantes = ifelse(categoria %in% c(10, 9, 15, 19, 35, 25, 32), categoria, estudantes)) %>% 
  mutate(estudantes = as.numeric(estudantes)) %>% 
  group_by(nome_programa) %>% 
  summarise(thesis = sum(estudantes))

# artigos
articles <- formulario %>% 
  dplyr::filter(PELD == "Sim") %>% 
  dplyr::select(nome_programa, producao_n) %>% 
  distinct() %>% 
  rename(articles = producao_n)

# outro produto
other_p <- formulario %>% 
  dplyr::filter(PELD == "Sim") %>% 
  dplyr::select(nome_programa, outro_produto) %>% 
  unique() %>%
  mutate(outro_produto = strsplit(as.character(outro_produto), ";")) %>%
  unnest(cols = c(outro_produto)) %>% 
  mutate(outro_produto = trimws(.$outro_produto)) %>% 
  separate(outro_produto, c("categoria", "quant"), " = ") %>% 
  mutate(quant = as.numeric(quant) %>% 
           replace_na(., 1),
         categoria = plyr::mapvalues(categoria, from = c("capitulo_livro", "livros", "protocolos", "videos", 
                                                         "trabalhos_tecnicos", "jogos", "desenho animado Mar a Vista",
                                                         "Manual de Monitoramento Reef Check Brasil", "Videos",
                                                         "livros_capitulos", "documentario"),
                                     to = c("book_chapter", "book", "protocol", "video", 
                                            "protocol", "game", "cartoon",
                                            "protocol", "video", "book_chapter", "video"))) %>% 
  pivot_wider(names_from = categoria, values_from = quant) %>% 
  data.frame() %>% 
  replace(., is.na(.), 0)

# join summaries
left_join(students, thesis) %>% 
  left_join(articles) %>% 
  left_join(other_p) %>% 
  knitr::kable(., caption = "Table 1. Scientific and social production of PELD sites research groups")

```


Most of the monitoring sites were relatively young and had continuous series < 4 years long, with the longest time associated with the PELD Lagoa dos Patos (ELPA) since 1993, but with the largest series of data associated with the monitoring carried out by the Brazilian Navy at the Instituto de Estudos do Mar Almirante Paulo Moreira (IEAPM) in Arraial do Cabo (since 1974). The majority (79%) of the sites had an annual monitoring frequency since the start of programs (figure 1, table 1), although some had a greater effort during some period (e.g., weekly, quarterly or monthly during a couple of years. Considering the PELD stablished periods, the plankton-associated groups had the longest time of monitoring, especially associated with the oldest coastal PELD sites (ELPA and PEBG, Figure 1). Elements describing geochemical processes and multitaxa approaches were part only of recently added site to the PELD program (HCES).
	The publication of manuscripts in scientific journals, book and book chapters constitute a large part of the results of those monitoring efforts and greatly increase (exponentially) together with the program number of years (figure 2). Also, the formation and training of scientists is valuable outcome of the programs, having accounting for 254 PhD and master’s thesis and 612 scientific papers in < 30 years.


```{r prod}

triste <- dados %>% 
  filter(peld != "non-PELD") %>% 
  group_by(nome_programa) %>% 
  summarise(ano = min(peld_since)) %>% 
  left_join(bind_cols(students, thesis[-1])) %>% 
  left_join(articles) %>% 
  left_join(other_p) %>% 
  mutate_each_(funs(as.numeric(.)), c("book_chapter", "book", "articles")) %>% 
  mutate(duration = 2021 - ano,
         all = book + book_chapter + articles + thesis) %>% 
  mutate(book = (book_chapter + book) / all,
         art = as.numeric(articles) / all,
         thes = as.numeric(thesis) / all) %>% 
  select(nome_programa, duration, all, book, art, thes)
  
triste[7,4] <- 1 # JAG, book of local species Arraial do Cabo

# https://cran.r-project.org/web/packages/scatterpie/vignettes/scatterpie.html#scatter-pie-plot

a <- ggplot(data = triste, aes(x=duration, y=all)) + 
  geom_point() +
  geom_smooth(method = "glm", formula = y~x,
              method.args = list(family = gaussian(link = 'log')))  +
  #coord_equal() + 
  theme_classic() +
  theme(legend.title = element_blank()) + 
  geom_text(aes(label = nome_programa))


b <- ggplot(data = triste, aes(x=duration, y=all)) + 
  geom_scatterpie(data = triste, aes(x=duration, y=log(all), group=nome_programa, r = 2), 
                  cols=c("book", "art", "thes")) +
  coord_equal() + 
  theme_classic() +
  theme(legend.title = element_blank())

a / b

# equation
glm(all ~ duration, data = triste, family = 'poisson') %>% summary()

```
Figure 2. Relationship between scientific production and age of long-term monitoring sites (PELD) in Brazilian coastal ecosystems. Generalized Linear Model adjusted to values by using Poisson distribution and log as link function (p < 0.001, AIC = 386.7).


```{r fund_outreach, echo=FALSE}

fund <- formulario %>% 
  dplyr::filter(PELD == "Sim") %>% 
  dplyr::select(nome_programa, fontes:aporte_valor) %>% 
  mutate(fontes1 = ifelse(!fontes %in% c("CAPES", "CNPq", "empresas", "FUNBIO", "pesquisador"), "FAPs", fontes)) %>%
  mutate(fontes1 = plyr::mapvalues(fontes1, from = c("empresas", "pesquisador"),
                                   to = c("private", "self funded")),
         nome_programa = gsub("PELD-", "", nome_programa)) %>% 
  ggplot(aes(x=nome_programa, y=aporte_valor/1000, fill=fontes1)) +
  geom_col() +
  theme_classic() +
  theme(legend.title = element_blank(),
        axis.text.x=element_text(angle=90, vjust = 0.5, hjust = 1),
        legend.position = c(0.85, 0.8)) +
  labs(x = "", y = "Funding (BRL*1000)", title = "(A)") +
  scale_x_discrete(limits = c("ELPA", "RLaC", "CCAL", "TAMS", "ILOC", "HCES", "RECA"))


outreach <- formulario %>% 
  dplyr::filter(PELD == "Sim") %>% 
  dplyr::select(nome_programa, divulgacao_eventos:divulgacao_outro) %>% 
  pivot_longer(cols = -nome_programa) %>% 
  distinct() %>% 
  mutate(name = gsub("divulgacao_", "", name),
         value = plyr::mapvalues(value, from = c("Esporadicamente", "Nunca", "Sempre"),
                                 to = c("sometimes", "never", "always"))) %>% 
  mutate(name = plyr::mapvalues(name, from = c("eventos", "jormal", "website", "midia_social", "radio", "tv",  "outro"),
                                to = c("events", "newspaper", "website", "social media", "radio", "tv",  "other"))) %>% 
  filter(!is.na(value)) %>%
  ggplot(aes(x=name, fill=value)) +
    geom_bar() +
    theme_classic() +
    theme(#axis.text.y=element_blank(),
      axis.text.x=element_text(angle=90, vjust = 0.5, hjust = 1, size = 12),
      legend.title = element_blank()) +
    labs(x = "", y ="", title = "(B)") + 
    scale_x_discrete(limits = c("events", "website", "social media", "tv", "newspaper", "radio", "other"))

fund + outreach

```
Figure 3. Accumulated funding received (A) and medias used to communicate monitoring outcomes (B) of long-term monitoring sites (PELD) in Brazil.


The CNPq was the main funding source of PELD sites, followed by regional funding agencies (FAP’s) and the CAPES (Coordenação de Aperfeiçoamento de Pessoal de Nível Superior), this last, usually supporting programs by supplying PhD and master’s scholarships (figure 3). Apart from scientific peer reviewed publications, most of results from PELD sites has been shown on scientific events and websites, with decreasing participation of large-scale medias such as broadcast television, newspaper and radio (figure 3).
PELD sites, although located along the coast and sharing same target habitats had very low connectivity with a maximum of two nodes shared between PELD nuclei. Only one site was isolated from all others (CCAL), but RECA and ELPA were weakly linked the rest of the network (figure 4). The participation of international institutions (19% of all nodes) was limited and associated mainly with the RLaC nucleus. NGOs represented only 3% (one node) of all connections and were located at the isolated group from CCAL.
The governmental institutes participation was present at most groups, but still poorly represented (10% of nodes) in observed networks.


```{r network, echo=FALSE}

# institutions' network

links <- formulario %>% 
  dplyr::filter(PELD == "Sim") %>% 
  filter(!is.na(pesquisador_responsavel)) %>% 
  dplyr::select(nome_programa, outras_instituicoes) %>% 
  unique()

links1 <- links %>%
  mutate(parcerias = strsplit(as.character(outras_instituicoes), ";")) %>%
  unnest(cols = c(parcerias)) %>%
  filter(parcerias != "") %>%
  select(-outras_instituicoes) %>% 
  mutate(parcerias = trimws(.$parcerias)) %>% 
  # bind_rows(data.frame(nome_programa = c("PELD-ILOC", "PELD-ELPA"), parcerias = c("UFF", "FURG"))) %>% 
  mutate(nome_programa = gsub("PELD-", "", nome_programa)) %>% 
  arrange(nome_programa) %>% 
  data.frame() %>% 
  bind_rows(data.frame(nome_programa = c("HCES", "HCES", "RECA", "RECA", "RLaC", "ILOC", rep("ELPA", 19)),
                       parcerias = c("Oregon State University", "UFF", "IEAPM", "ICMBio", "UFRJ", "UFF",
                                     "FURG", "UFRGS", "UFPR", "UFC", "UFPE", "UERJ", "Texas A&M University", 
                                     "University of North Texas", "The University of Adelaide", "NOAA",
                                     "Portos RS", "ICMBio", "MCTI", "MAPA", "Yaqu Pacha", "NEMA", 
                                     "KAOSA", "Costa Semi-Árida", "TAMS")))

# network
net <- graph.data.frame(links1, directed=T) %>% 
  simplify(., remove.multiple = T, remove.loops = T)
# E(net)$arrow.size <- .1
# plot(net)

# get nodes
plotcord <- data.frame(layout.fruchterman.reingold(net))
plotcord$subjects <- c(unique(links1 %>% pull(nome_programa)), unique(links1 %>% pull(parcerias))) %>% unique()

plotcord <- plotcord %>% 
  mutate(tamanho = ifelse(subjects %in% c("Abrolhos", "CCAL", "ELPA", "HCES", "ILOC", "PEBG", "RECA", "RLaC", "TAMS"), 3, 1),
         cor_node = ifelse(subjects %in% c("NEMA", "MCTI", "MAPA", "Portos RS", "ICMBio", "Embrapa", "CEPENE", "Ibama", "INPE", "Jardim Botânico do Rio de Janeiro", "IEAPM"), "govermental institution",
                           ifelse(subjects %in% c("Oregon State University", "Goethe Universität Frankfurt/Alemanha", 
                                                  "University of Kiel/Alemanha", "Universidad Nacional de Mar del Plata/Argentina",
                                                  "Universidad de la Republica/Uruguai", "Environment Agency/Austria", 
                                                  "University of Southern California", "University of South Carolina/USA",
                                                  "Oregon State University","Texas A&M University", "NOAA",
                                                  "University of North Texas", "The University of Adelaide"), "foreign institution",
                                  ifelse(subjects %in% c("Abrolhos", "CCAL", "ELPA", "HCES", "ILOC", "PEBG", "RECA", "RLaC", "TAMS"), "PELD project",
                                         ifelse(subjects %in% c("ONG Biota", "KAOSA", "Yaqu Pacha"), "NGO", 
                                                ifelse(subjects %in% c("Costa Semi-Árida", "TAMS"), "PELD project", "Brazilian university")))))
  )


# get vertices
g <- get.data.frame(net)  # get the edge information using the get.data.frame function


# add codes
plotcord <- left_join(plotcord,
                      plotcord %>% 
                        select(-X1,-X2,-tamanho) %>% 
                        filter(!subjects %in% unique(g$from)) %>% 
                        mutate(codes = ifelse(subjects %in% unique(g$from), subjects,
                                              as.factor(subjects) %>% as.numeric())) %>% 
                        mutate(codes = as.numeric(codes))) %>% 
  arrange(codes) %>% 
  mutate(codes = ifelse(subjects %in% unique(g$from), subjects, codes))


# add vertices
g$from.x <- plotcord$X1[match(g$from, plotcord$subjects)]  #  match the from locations from the node data.frame we previously connected
g$from.y <- plotcord$X2[match(g$from, plotcord$subjects)]
g$to.x <- plotcord$X1[match(g$to, plotcord$subjects)]  #  match the to locations from the node data.frame we previously connected
g$to.y <- plotcord$X2[match(g$to, plotcord$subjects)]

# plot
g %>% 
  ggplot() +
  geom_segment(alpha = 1, aes(x = from.x, xend = to.x, y = from.y, yend = to.y)) +
  geom_point(data = plotcord, 
             aes(x = X1, y = X2, colour = cor_node, size = tamanho)) +
  geom_text_repel(data = plotcord, 
                  aes(x = X1, y = X2, label = codes),
                  box.padding = 0.3) +
  theme_classic() +
  theme(legend.title = element_blank()) +
  # lims(x = c(-10, 15)) +
  scale_size(range = c(1, 5)) 

```

Table 2s. Acronyms of institutions involved with marine and coastal long-term monitoring sites (PELD) in Brazil from figure 4.  
```{r}

# list of acronyms for network plot
plotcord %>% 
  select(-X1, -X2, -tamanho) %>% 
  filter(cor_node != "PELD project") %>% 
  mutate(codes = as.numeric(codes)) %>% 
  left_join(links1 %>% 
              rename(subjects = parcerias)) %>% 
  arrange(codes, nome_programa) %>% 
  mutate(n = "amod") %>% 
  pivot_wider(
    names_from = "n", 
    values_from = "nome_programa"
  ) %>% 
  knitr::kable()

# write.csv(g, "data/g_network_pelds.csv", row.names = F)
# write.csv(plotcord, "data/plotcord_network_pelds.csv", row.names = F)

```


```{r}

# network metrics

# https://kateto.net/networks-r-igraph
edge_density(net, loops=F) # density of a graph is the ratio of the number of edges and the number of possible edges.

# calculates the average path length in a graph, by calculating the shortest paths between all pairs of vertices
mean_distance(net, directed=F) 
centr_clo(net, mode="all", normalized=T) 
modularity(ceb)

```

## Discussion (1000 words)

Long-term monitoring projects in the marine environment in Brazil were found to be young and no well-connected, but scientifically productive with large importance for academic and human resources training. Long-term ecological monitoring programs are sought to give the planned return after several years of existence (Caughlan & Oakley, 2001)  because the questions raised by this model of ecological approach need long time series to allow depicting temporal fluctuation patterns (Giron-Nava et al., 2017). Notwithstanding, date presented here indicates that the scientific products grow exponentially with the programs’ age, which was also observed for terrestrial PELD sites (Randig 2019). Thus, it is clear that efforts and investment in such programs have considerable contribution to scientific advances besides leaving a legacy on human resources and networking which also boosts other outcomes such as policy making (Hughes et al., 2017). Unfortunately, socio-economic impacts of PELDs were not evaluated here because of the complexity of potential indicators. 
	Changes in global environmental stability and its effects over biodiversity and ecosystems services sustainability are main concerns of global policy (Haase et al., 2018) and the monitoring of changes is a key point. Monitoring of essential biological and oceanographic variables, EBVs and EOVs sequentially, have been one of the aims of global strategies focused on United Nations Sustainable Development Goals (Kissling et al., 2018; Miloslavich et al., 2018). In this sense, long-term monitoring programs are an essential tool to detect changes associated to anthropogenic stressors based on those essential variables separating real effects from background noise (Magurran et al., 2010). Variables and biological groups monitored by the Brazilian PELD programs are within those that are EOVs and EBVs, such as phytoplankton/zooplankton abundance, macroalgae cover and fish abundance (Miloslavich et al., 2018). Despite most of PELD programs being relatively young, the monitoring of those key groups at large spatial scale is advantageous for complying with emerging international efforts.
	PELD sites included most of coastal environments along the Brazilian coast although monitoring points were scattered among habitats. Asymmetries among monitored variables and habitats found indicate older and more diverse initiatives concentrated at southern sites in the Brazilian coast where oldest universities and institute were established. Considering the habitats covered there also gaps related to the representativeness of estuaries and intertidal areas in general, but especially at the north and northeast region, which leave aside the largest portion of those habitats along the coast (Diniz et al., 2019). The continental dimensions of the country are a natural challenge for monitoring the diversity of habitats on equal manner. Reef associated organisms and the ecosystem itself were better covered by monitoring initiatives, but important offshore and non-consolidate habitats are absent and underrepresented, respectively. Magris et al. (2021) indicated a priority maps of marine habitats according to multiple threats, habitat representativeness, endangered species and connectivity. All of the marine and coastal PELD sites lies within or in the vicinities of those priority areas (figure 1), with PEBG and Abrolhos representing top-priority areas. The indication of poor representation and large gaps on monitoring along priority areas raises a red flag for potential targets on the establishment of new monitoring sites. All programs had monitoring points within marine or coastal protected areas which usually had historical relationships with the MPA creation or management demand. Focusing the monitoring of ecosystems on protected area may bias the knowledge about important and greatly threaten area located close to populated areas. Only the PEGB was found to be located exactly in an important mosaic of ecosystem that surrounds the most densely populated city in the Brazilian coast (i.e., Rio de Janeiro city, IBGE 2018). 
The actual scenario indicates the need of expanding the monitoring sites. However, long-term monitoring initiatives rely on networks of scientists and institutions to support the continuous and standardized effort of sampling (Magurran et al., 2010). Another crucial aspect indicated by coordinators is related to financial support of those initiatives, that usually went through variable periods of monitoring effort where core activities were preserved during period of funds shortage, to expanding in groups and variables sampled during less troubled periods (Caughlan & Oakley, 2001).
	Continuous financing support is a crucial matter for long-term ecological monitoring programs (Caughlan & Oakley, 2001) and, indeed, it was the main difficult of cited by all programs and initiatives present at the workshop. We had not included data about the individual monitoring initiatives but it is worth noting that those were supported by funding granted to involved coordinators either by small grants or self-funding that kept those monitoring as side project because of their ecological importance. Concerning the PELD programs, prospecting other funds to complement or expanding activities was a regular task. The PELD program supported by CNPq is <25 years-old, but has had supported several initiatives in key ecosystems of Brazil, mainly on the continent, despite their shortage of resources. Scientific institutions had been continuously depleted of funds in last years and the future scenario is a blurred picture and long-term monitoring programs are at risk (de Oliveira Andrade, 2019; Fernandes et al., 2017; Hallal, 2021; Magnusson et al., 2018). The integration of efforts with other sources of financing such as from funding managers (e.g., FUNBIO) and from national mixed-sourced grantees (e.g., Instituto Serrapilheira, Fundação ‘O Boticário’) may help increase the life spam of monitoring programs. Private sector funded monitoring programs in Brazil are mostly species or taxa focused (e.g., sea turtles, marine mammals, corals) probably because it is easily sold to society, but that view restricts the application of ecosystem-level and multitaxa monitoring programs in grants calls.
	Strengthen long-monitoring ecological initiatives may contribute to increase the odds of consortium type applications for international grants allowing also widening the scale of questions answered by those programs (Haase et al., 2018). The Ocean Decade and the integration of several monitoring programs, e.g. ILTER and GEO BON (Muelbert et al., 2019), may facilitate the integration of working monitoring which increases the possibilities of access to new funds. Part of integration efforts should aim to standardize monitoring protocols and leveraging human resources formation by providing access to training in best practices and updating of methods. Lastly, increasing the communication among scientists and monitoring programs is important and will help to sustain and expand monitoring initiatives, but the delivery of results and the appropriation of the generated knowledge from those initiatives to the general public and decision makers may be the most important aim of all. ‘The science we need for the ocean we want’ is the Ocean Decade’s motto, and is a general feeling that must be taken seriously for those hoping for a sustainable ocean in which scientists keep monitoring for centuries to come.


## References (40 refs)
Brito, M. A., Oliveira, D. De, & Mamede, M. D. A. (2020). Programa de Pesquisa Ecológica de Longa Duração – PELD / CNPq: desafios da gestão, avanços e perspectivas. Oecologia Australis, 24(2), 259–265.

Caughlan, L., & Oakley, K. L. (2001). Cost considerations for long-term ecological monitoring. Ecological Indicators, 1(2), 123–134. https://doi.org/10.1016/s1470-160x(01)00015-2

de Oliveira Andrade, R. (2019). Brazil’s budget cuts threaten more than 80,000 science scholarships. Nature, 572(7771), 575–576. https://doi.org/10.1038/d41586-019-02484-w

Diniz, C., Cortinhas, L., Nerino, G., Rodrigues, J., Sadeck, L., Adami, M., & Souza-Filho, P. W. M. (2019). Brazilian mangrove status: Three decades of satellite data analysis. Remote Sensing, 11(7). https://doi.org/10.3390/rs11070808

Fernandes, G. W., Vale, M. M., Overbeck, G. E., Bustamante, M. M. C., Grelle, C. E. V., Bergallo, H. G., … Pillar, V. D. (2017). Dismantling Brazil’s science threatens global biodiversity heritage. Perspectives in Ecology and Conservation, 15(3), 239–243. https://doi.org/10.1016/j.pecon.2017.07.004

Gaiser, E. E., Bell, D. M., Castorani, M. C. N., Childers, D. L., Groffman, P. M., Jackson, C. R., … Zinnert, J. C. (2020). Long-Term Ecological Research and Evolving Frameworks of Disturbance Ecology. BioScience, 70(2), 141–156. https://doi.org/10.1093/biosci/biz162

Giron-Nava, A., James, C. C., Johnson, A. F., Dannecker, D., Kolody, B., Lee, A., … Sugihara, G. (2017). Quantitative argument for long-term ecological monitoring. Marine Ecology Progress Series, 572, 269–274. https://doi.org/10.3354/meps12149

Haase, P., Tonkin, J. D., Stoll, S., Burkhard, B., Frenzel, M., Geijzendorffer, I. R., … Schmeller, D. S. (2018). The next generation of site-based long-term ecological monitoring: Linking essential biodiversity variables and ecosystem integrity. Science of the Total Environment, 613–614, 1376–1384. https://doi.org/10.1016/j.scitotenv.2017.08.111

Hallal, P. C. (2021). SOS Brazil: science under attack. The Lancet, 397(10272), 373–374. https://doi.org/10.1016/S0140-6736(21)00141-0

Heymans, J. J., Bundy, A., Christensen, V., Coll, M., de Mutsert, K., Fulton, E. A., … Travers-Trolet, M. (2020). The Ocean Decade: A True Ecosystem Modeling Challenge. Frontiers in Marine Science, 7(September 2020), 1–5. https://doi.org/10.3389/fmars.2020.554573

Hughes, B. B., Beas-Luna, R., Barner, A. K., Brewitt, K., Brumbaugh, D. R., Cerny-Chipman, E. B., … Carr, M. H. (2017). Long-Term studies contribute disproportionately to ecology and policy. BioScience, 67(3), 271–278. https://doi.org/10.1093/biosci/biw185

Kissling, W. D., Ahumada, J. A., Bowser, A., Fernandez, M., Fernández, N., García, E. A., … Hardisty, A. R. (2018). Building essential biodiversity variables (EBVs) of species distribution and abundance at a global scale. Biological Reviews, 93(1), 600–625. https://doi.org/10.1111/brv.12359

Lindenmayer, D. B., & Likens, G. E. (2009). Adaptive monitoring: a new paradigm for long-term research and monitoring. Trends in Ecology and Evolution, 24(9), 482–486. https://doi.org/10.1016/j.tree.2009.03.005

Magnusson, W. E., Braga-Neto, R., Flávia Pezzini, Baccaro, F., Bergallo, H., Penha, J., … Pontes, A. R. M. (2013). Biodiversity and Integrated Environmental Monitoring: a case for the RAPELD method. Manaus: Attema.

Magnusson, W. E., Grelle, C. E. V., Marques, M. C. M., Rocha, C. F. D., Dias, B., Fontana, C. S., … Fernandes, G. W. (2018). Effects of Brazil’s political crisis on the science needed for biodiversity conservation. Frontiers in Ecology and Evolution, 6(OCT), 1–5. https://doi.org/10.3389/fevo.2018.00163

Magurran, A. E., Baillie, S. R., Buckland, S. T., Dick, J. M. P., Elston, D. A., Scott, E. M., … Watt, A. D. (2010). Long-term datasets in biodiversity research and monitoring: Assessing change in ecological communities through time. Trends in Ecology and Evolution, 25(10), 574–582. https://doi.org/10.1016/j.tree.2010.06.016

Miloslavich, P., Bax, N. J., Simmons, S. E., Klein, E., Appeltans, W., Aburto-Oropeza, O., … Shin, Y. J. (2018). Essential ocean variables for global sustained observations of biodiversity and ecosystem changes. Global Change Biology, 24(6), 2416–2433. https://doi.org/10.1111/gcb.14108

Mirtl, M., T. Borer, E., Djukic, I., Forsius, M., Haubold, H., Hugo, W., … Haase, P. (2018). Genesis, goals and achievements of Long-Term Ecological Research at the global scale: A critical review of ILTER and future directions. Science of the Total Environment, 626(December 2017), 1439–1462. https://doi.org/10.1016/j.scitotenv.2017.12.001

Muelbert, J. H., Nidzieko, N. J., Acosta, A. T. R., Beaulieu, S. E., Bernardino, A. F., Boikova, E., … Zingone, A. (2019). Ilter – the international long-term ecological research network as a platform for global coastal and ocean observation. Frontiers in Marine Science, 6(AUG), 1–14. https://doi.org/10.3389/fmars.2019.00527

Sheldon, K. S. (2019). Climate Change in the Tropics: Ecological and Evolutionary Responses at Low Latitudes. Annual Review of Ecology, Evolution, and Systematics, 50, 303–333. https://doi.org/10.1146/annurev-ecolsys-110218-025005

Visbeck, M. (2018). Ocean science research is key for a sustainable future. Nature Communications, 9(1), 1–4. https://doi.org/10.1038/s41467-018-03158-3

Worm, B., & Lotze, H. K. (2016). Marine Biodiversity and Climate Change. Climate Change: Observed Impacts on Planet Earth: Second Edition (Second Edi). Elsevier B.V. https://doi.org/10.1016/B978-0-444-63524-2.00013-0

Barbosa FAR (2013) Uma breve história do Programa de Pesquisas Ecológicas de Longa Duração (PELD – CNPq) do Brasil: da semente ao fruto. In: Tabarelli M, Rocha CFD, Romanowski HP, Rocha O, Lacerda LD (eds) PELD – CNPq: dez anos do Programa de Pesquisas Ecológicas de Longa Duração no Brasil: achados, lições e perspectivas. Editora Universitária UFPE, Recife, Brazil, 446p. Available in Portuguese at https://www.lume.ufrgs.br/bitstream/handle/10183/104869/000937661.pdf?sequence=1. Accessed in 8th May 2021.

IBGE, 2018. Instituto Brasileiro de Geografia e Estatística. Características da população e dos domicílios: resultados do universo. Available in Portuguese at https://cidades.ibge.gov.br/brasil/. Retrieved June 17, 2018. Accessed in 8th May 2021.

Randig O (2019) Aprimoramento do sistema de acompanhamento e avaliação do Programa de Pesquisa Ecológica de Longa Duração – PELD no âmbito do Conselho Nacional de Desenvolvimento Científico e Tecnológico – CNPq. Honors thesis, Escola Nacional de Administração Pública (Enap), Brasília/DF, Brazil. 21p. Available in Portuguese at: http://repositorio.enap.gov.br/handle/1/3895. Accessed in 8th May 2021.

